---
title: "p8105_hw2_md4270"
author: "Manye Dong"
date: "2023-09-30"
output: github_document
---
```{r message=FALSE, results='hide'}
library(tidyverse)
library(readr)
library(readxl)
```

## Problem 1
### clean pols_month dataset
```{r message=FALSE}
# clean snp dataset
pols_month = 
  read_csv("pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, sep = "-", into = c("year", "month", "day")) |>
  mutate(year = as.numeric(year), month = as.numeric(month), 
         day = as.numeric(day), month = month.name[month],
         president = ifelse(prez_gop==0, "dem", "gop")) |>
  select(-c(prez_dem, prez_gop, day))

head(pols_month)
```

### clean snp dataset
```{r message=FALSE}
snp = 
  read_csv("snp.csv") |>
  janitor::clean_names() |>
  separate(date, sep = "/", into = c("month", "day", "year")) |>
  mutate(year = as.numeric(year), month = as.numeric(month), 
         day = as.numeric(day), month = month.name[month], 
         year = ifelse((50<=year & year<100), 1900+year, 2000+year)) |>
  select(year, month, close)

head(snp)
```
### Clean Unemployment Dataset
```{r message=FALSE}
unemployment = 
  read_csv("unemployment.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    jan:dec,
    names_to = "month", 
    values_to = "umployment_percent") |>
  mutate(year = as.numeric(year), 
         month = recode(month, 
                   "jan" = "January", 
                   "feb" = "February", 
                   "mar" = "March", 
                   "apr" = "April", 
                   "may" = "May", 
                   "jun" = "June", 
                   "jul" = "July", 
                   "aug" = "August", 
                   "sep" = "September", 
                   "oct" = "October", 
                   "nov" = "November", 
                   "dec" = "December"))

head(unemployment)
```
```{r}
# merge the three datasets
joined = left_join(pols_month, snp) |>
  left_join(x = _, y = unemployment)
joined
```
Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables):

The resulting dataset contains `r nrow(joined)` rows and `r ncol(joined)` columns. The range of the year goes from `r min(joined %>% pull(year))` to `r max(joined %>% pull(year))`.


## Problem 2
### Mr. Trash Wheel
```{r}
mister_trash = 
  read_excel("202309 Trash Wheel Collection Data.xlsx", 
             sheet = "Mr. Trash Wheel", range = "A2:N586") |>
  janitor::clean_names() |>
  separate(date, sep ="-", into = c("year", "month", "day")) |>
  mutate(homes_powered = (weight_tons * 500) / 30, 
         month = month.abb[as.numeric(month)],
         wheel = "mister")

tail(mister_trash)
```
```{r}
# load and clean the Professor Trash Wheel
prof_trash = 
  read_excel("202309 Trash Wheel Collection Data.xlsx", 
             sheet = "Professor Trash Wheel", range = "A2:M108") |>
  janitor::clean_names() |>
  separate(date, sep ="-", into = c("year", "month", "day")) |>
  mutate(homes_powered = (weight_tons * 500) / 30, 
         month = month.abb[as.numeric(month)],
         wheel = "professor")

tail(prof_trash)
```
```{r}
# load and clean the Professor Trash Wheel
gwynnda_trash = 
  read_excel("202309 Trash Wheel Collection Data.xlsx", 
             sheet = "Gwynnda Trash Wheel", range = "A2:L157") |>
  janitor::clean_names() |>
  separate(date, sep ="-", into = c("year", "month", "day")) |>
  mutate(homes_powered = (weight_tons * 500) / 30, 
         month = month.abb[as.numeric(month)],
         wheel = "gwynnda")

tail(gwynnda_trash)
```
Now we combine those three datasets:
```{r}
trash_wheels =
  bind_rows(mister_trash, prof_trash, gwynnda_trash) |>
  janitor::clean_names()

tail(trash_wheels)
```
* There are `r nrow(trash_wheels)` rows/observations in the resulting dataset. 

* Key variables include dumpster id, year-month-day as the time of the record for the 

* The total weight of trash collected by Professor Trash Wheel is `r sum(prof_trash %>% pull(weight_tons))` in tons.

```{r results='hide'}
filtered_gywnnda =  
  gwynnda_trash |>
  select(year, month, cigarette_butts) |>
  filter(year=="2021", month=="Jul")

head(filtered_gywnnda)
```
* The total number of cigarette butts collected by Gwynnda in July of 2021 is `r sum(filtered_gywnnda$cigarette_butts)` (16300).


## Problem 3
### Baseline Dataset
```{r message=FALSE}
MCI_baseline = read_csv("MCI_baseline.csv", skip=1)
```
```{r warning=FALSE}
MCI_cleaned = 
  MCI_baseline |>
  janitor::clean_names() |>
  mutate(sex = ifelse(sex==0, "Female", "Male"), 
         apoe4 = ifelse(apoe4==0, "Non-carrier", "Carrier"),
         age_at_onset = as.numeric(age_at_onset)) |>
  filter(age_at_onset != "NA")

head(MCI_cleaned)
```

* I used "skip=1" parameter when importing data to exclude the first row of the because it will mess up with the column names one row below. 

* Relevant features of the dataset include current age of participants, their gender, education in years, and the age at onset of MCI in decimal form after data cleaning.

* There were originally `r nrow(MCI_baseline)` participants recruited, of these `r nrow(MCI_cleaned)` develop MCI the follow-up period. 

* The average baseline age is `r mean(MCI_cleaned %>% pull(age_at_onset))`.

```{r}
# Check the proportion of women who are APOE4 carriers
MCI_women = 
  MCI_cleaned |>
  filter(sex=="Female")

MCI_women_carrier = 
  MCI_cleaned |>
  filter(sex=="Female", apoe4=="Carrier")
```

* `r 100 * nrow(MCI_women_carrier) / nrow(MCI_women)`% of women in the study are APOE4 carriers.

### Longitudinal Study
```{r}
amyloid =
  read_csv("mci_amyloid.csv", skip = 1) |>
  janitor::clean_names()

head(amyloid)
```
Here, I also used "skip=1" parameter to exclude the first row because it is not the column name we want. Key features include 

Now we check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. 
```{r}
baseline_id = MCI_cleaned |> pull(id)
amyloid_id = amyloid |> pull(study_id)
```
So, yes, there are many participants who appear in only the amyloid dataset, and a lot only appear in the baseline dataset.

Now, we combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained:
```{r}
MCI_df = inner_join(MCI_cleaned, amyloid, by = c("id" = "study_id"))

head(MCI_df)
```
The resulting dataset contains `r nrow(MCI_df)` rows/observations and `r ncol(MCI_df)` columns. Key features include the study id for each participant, their current age, gender, education level in years and their age at the onset of MCI. The rest columns are measurements of biomarkers amyloid, from baseline level to the levels after different time intervals across the study.

Finally, we export the result as a CSV to current data directory:
```{r save_csv, message = FALSE}
write.csv(MCI_df, file = "MCI_df.csv")
```
